---
title: "SDMsel"
subtitle: "Species Distribution Model Selection"
bibliography: library.bib
output:
  html_document:
    theme: flatly
    highlight: default
    toc: true
    toc_depth: 2
    toc_float: true
vignette: >
  %\VignetteIndexEntry{SDMsel}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(knitr.table.format = "html")
```

# Set working environment

```{r kableExtra, echo=F}
library(knitr)
```

`SDMsel` (Species Distribution Model Selection) package works with the new version of **MaxEnt** software (v. >= 3.4.1), in case you have the old version you can download the new one from [here](https://biodiversityinformatics.amnh.org/open_source/maxent/). The new version uses by default the **complementary log-log transform** (Cloglog) to produce the estimate of occurrence probability instead of the **logistic transform** and the feature class **threshold** is disabled but available as option [@Phillips2017].

If is the first time you use **MaxEnt** in R you have to make available the **maxent.jar** file for the `dismo` package [@Hijmans2017]. To do that copy the **maxent.jar** file in the folder named **java** inside the folder returned by the following command:

```{r, eval=FALSE}
system.file(package="dismo")
```

If you are an Ubuntu user you have to install **java** (e.g. install java openjdk with `sudo apt install openjdk-11-jdk`), check it with `java --version`, and then configure java for `R` using the command `sudo R CMD javareconf`.

Before loading `SDMsel` you have to increase the RAM that **MaxEnt** can use. You should leave around 2G for your OS and make the rest available for MaxEnt (e.g. if you have 4GB RAM make 2G available using `-Xmx2g`): 

```{r java option}
options(java.parameters = "-Xmx2g" )
```

Import `SDMsel`:
```{r import}
library(SDMsel)
```

Check if maxent is correctly configured for `dismo`:

```{r check maxent}
dismo::maxent()
```

Now everything is set up and you can start your analysis. A good starting point to learn how to use **SDMsel** is the article [Tune model using AUC]() in the package website.

(**TODO: I will add the right link when we release the pkg!**)

# References
