---
title: "Species Distribution Model Selection"
author: "Sergio Vignali, Arnaud Barras, Veronika Braunisch"
date: "`r Sys.Date()`"
bibliography: library.bib
output: 
  prettydoc::html_pretty:
    df_print: kable
    theme: cayman
    highlight: github
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Maxent Model Selection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(knitr.table.format = "html")
```

# Set working environment

```{r kableExtra, echo=F}
library(knitr)
library(kableExtra)
```

`SDMsel` (Species Distribution Model Selection) package works with the new version of **MaxEnt** software (v. >= 3.4.1), in case you have the old version you can download the new one from [here](https://biodiversityinformatics.amnh.org/open_source/maxent/). The new version uses by default the **complementary log-log transform** (Cloglog) to produce the estimate of occurrence probability instead of the **logistig transform** and the feature class **threshold** is disabled but available as option [@Phillips2017].

If is the first time you use **MaxEnt** in R you have to make available the **maxent.jar** file for the `dismo` package [@Hijmans2017]. To do that copy the **maxent.jar** file in the folder named **java** inside the folder returned by the following command (only for Ubuntu, install java openjdk witn `sudo apt install openjdk-11-jdk`, check it with `java --version`, and after run `sudo R CMD javareconf`):

```{r, eval=FALSE}
system.file(package="dismo")
```

Before loading `SDMsel` you have to increase the RAM that **MaxEnt** can use. You should leave around 2G for your OS and make the rest available for MaxEnt (e.g. if you have 4GB RAM make 2G available using `-Xmx2g`): 

```{r java option}
options(java.parameters = "-Xmx2g" )
```

Import `SDMsel`:
```{r import}
library(SDMsel)
```

Load packages used in the vignette:

```{r}
library(dismo)      # To download data used in the vignette
library(ggplot2)    # To plot graphs
library(maps)       # For access useful maps
library(rasterVis)  # To plot raster objects
library(gridExtra)  # To plot multiple ggplot object in one page
```

Check if maxent is correctly configured for `dismo`:

```{r check maxent}
maxent()
```
<div style="text-align: right"><a href="#top">Back to top</a></div>


# Acquire environmental variables and species distribution data

For the analysis we will use the climat data from [WorldClim](http://www.worldclim.org/) version 1.4 [@Hijmans2005] (you can download the new version 2.0 from the previous link) and the ecoregions from [WWF](https://www.worldwildlife.org/publications/terrestrial-ecoregions-of-the-world) [@Olson2001] included in the `dismo` package:

```{r get predictors, fig.align="center"}
files <- list.files(path = paste(system.file(package = 'dismo'), '/ex', sep = ''), pattern = 'grd', full.names = T)
predictors <- stack(files)
```

Plot **bio1** using [gplot](https://www.rdocumentation.org/packages/rasterVis/versions/0.43/topics/gplot-methods) function from `rasterVis` package:

```{r, fig.align="center"}
gplot(predictors$bio1) +
    geom_tile(aes(fill = value)) +
    coord_equal() +
    scale_fill_gradientn(colours = c("#2c7bb6", "#abd9e9", "#ffffbf", "#fdae61", "#d7191c"),
                         na.value = "transparent",
                         name = '°C x 10') +
    labs(title = 'Annual Mean Temperature',
         x = 'longitude',
         y = 'latitude') +
    scale_x_continuous(expand = c(0, 0)) +
    scale_y_continuous(expand = c(0, 0)) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5),
          axis.ticks.x = element_blank(),
          axis.ticks.y = element_blank())
```

There are nine environmental variables:

* Continuous variables
    + **bio1** Annual Mean Temperature
    + **bio5** Max Temperature of Warmest Month
    + **bio6** Min Temperature of Coldest Month
    + **bio7** Temperature Annual Range (bio5-bio6)
    + **bio8** Mean Temperature of Wettest Quarter
    + **bio12** Annual Precipitation
    + **bio16** Precipitation of Wettest Quarter
    + **bio17** Precipitation of Driest Quarter
* Categorical variables
    + **biome** Terrestrial Ecoregions of the World

For demostrating how to use `SDMsel` we download a dataset of [*Vultur gryphus*](https://en.wikipedia.org/wiki/Andean_condor) from  the Global Biodiversity Inventory Facility [GBIF](http://www.gbif.org/) using the [gbif](https://www.rdocumentation.org/packages/dismo/versions/1.1-4/topics/gbif) function of `dismo` package. We download data from the last 3 years. 
```{r get presence data}
data <- gbif(genus = 'Vultur', species = 'gryphus*', geo = TRUE, removeZeros = TRUE, args = 'year=2015,2017')
data <- data.frame(x = data$lon, y = data$lat)
clean_data <- data[!duplicated(data), ]  # Remove duplicate rows
clean_data <- clean_data[!is.na(clean_data$x) | !is.na(clean_data$y), ]  # Remove NA values
```
We remove all but one location per raster cell using the function `thinData`:
```{r}
clean_data <- thinData(clean_data, predictors)
```


```{r}
cat('Number of observations before cleaning data:', nrow(data),
    '\nNumber of obsevations after cleaning data  :', nrow(clean_data))
```


We downloaded `r nrow(data)` presence locations and after cleaning the duplicates and the NA we have `r nrow(clean_data)` locations for our analysis. We can plot the study area and the presence locations we downloaded from **GBIF**:

```{r, fig.align="center"}
ggplot(map_data('world'), aes(long, lat)) +
    geom_polygon(aes(group = group), fill = "grey95", color = "gray40", size = 0.2) +
    geom_jitter(data = clean_data, aes(x = x, y = y), color = 'red',
                alpha = 0.4, size = 1) +
    labs(x = "longitude", y = "latitude") +
    theme_minimal() +
    theme(legend.position = "none") +
    coord_fixed() +
    scale_x_continuous(limits = c(-125, -32)) +
    scale_y_continuous(limits = c(-56, 40))
```
In this tutorial we don't check the validity of the locations, instead we assume they are correct and we model the distribution only for demonstration purpose.
<div style="text-align: right"><a href="#top">Back to top</a></div>


# Model Selection

MaxEnt is a Machine Learning (ML) approach for Species Distribution Models and most of the ML methods require to define several **hyperparameters** before training the model. Hyperparameters are parameters whose values are set before the learning process begins and are not estimated from the data. Their best value is unknown for a given model but can be assessd using a tuning precedure. During the tuning procedure several models are trained varying the value of a hyperparameter and the model with the best performance is retained. The performance of a model can be evaluated using various metrics, `SDMsel` provides three: `AUC`, `TSS` (True Skill Statistic) and `AICc`.

## Prepare data for Model Selection

In this vignette we demonstrate how to perform model selction using the `AUC`. We create a random train/val/test datasets withholding a 20% sample for validation and a 20% for testing. We set the seed to obtain consistent results between different trail:

```{r split data in train/dev/test dataset}
# The %<-% operator from "zeallot"" package unpacks a list in variables
c(train_coords, val_coords, test_coords) %<-% trainValTest(clean_data, val = 0.2, test = 0.2, seed = 2530)
cat('Total presence locations  :', nrow(clean_data),
    '\nTotal train locations     :', nrow(train_coords),
    '\nTotal validation locations:', nrow(val_coords),
    '\nTotal test locations      :', nrow(test_coords))
```

Plot train and test datasets:

```{r, fig.align="center"}
base_plot <- ggplot(map_data('world'), aes(long, lat)) +
  geom_polygon(aes(group = group), fill = "grey95", color = "gray40", size = 0.2) +
  labs(x = "longitude", y = "latitude") +
  theme_minimal() +
  theme(legend.position = "none") +
  coord_fixed() +
  scale_x_continuous(limits = c(-125, -32)) +
  scale_y_continuous(limits = c(-56, 40)) +
  theme(plot.title = element_text(hjust = 0.5))

train_plot <- base_plot + geom_jitter(data = train_coords, aes(x = x, y = y), color = 'red', alpha = 0.4, size = 1) + labs(title = 'Train')
val_plot <- base_plot + geom_jitter(data = val_coords, aes(x = x, y = y), color = 'green', alpha = 0.4, size = 1) + labs(title = 'Validation')
test_plot <- base_plot + geom_jitter(data = test_coords, aes(x = x, y = y), color = 'blue', alpha = 0.4, size = 1) + labs(title = 'Test')

grid.arrange(train_plot, val_plot, test_plot, nrow = 1)
```

To extract the background locations we use the [randomPoints](https://www.rdocumentation.org/packages/dismo/versions/1.1-4/topics/randomPoints) function from `dismo` package.

```{r extract background locations}
set.seed(2530)
bg_coords <- randomPoints(predictors, 10000)
```

The environmental variables we downloaded have a coarse resolution and the function cannot extract more than `r nrow(bg_coords)` as you can see from the warning message.

Before running the model we have to prepare the data in the correct format. The `prepareSWD` function will create a `SWD` object that stores the species name, the coordinates of the locations and the value of the environmental variables at location points. The parameter `categoricals` indicates which environmental variables are categorical. In our example **biome** is categorical (you can pass a vector if you have more than one categorical environmental variables). The function extracts the value of the environmental variables for each locations and excludes those locations that have `NA` value for at least one environmental variable.

```{r prepare SWD datasets}
train <- prepareSWD(species = "Vultur gryphus", coords = train_coords, env = predictors, categoricals = "biome")
val <- prepareSWD(species = "Vultur gryphus", coords = val_coords, env = predictors, categoricals = "biome")
test <- prepareSWD(species = "Vultur gryphus", coords = test_coords, env = predictors, categoricals = "biome")
bg <- prepareSWD(species = "Vultur gryphus", coords = bg_coords, env = predictors, categoricals = "biome")
```

The warning message reports that in the background dataset there are 9 locations that are discarded.  
Let's have a look at the `SWD` object:

```{r show SWD object}
train
```

To see the data we run:

```{r show train data, eval=F, echo=T}
train@data
```

```{r make table, echo=F}
kable(train@data) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "responsive")) %>%
  scroll_box(height = "500px")
```

or access the coordinates with `train@coords` or the species with `train@species`.

```{r show train coords data, eval=F, echo=T}
train@coords
```

```{r, echo=F}
kable(train@coords) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "responsive")) %>%
  scroll_box(height = "500px")
```

We can save the `SWD` object in a **.csv** file to use it directly in the software **MaxEnt**.
```{r save SWD, eval=F, echo=T}
swd2csv(train, file_name = "train.csv")
```

We use a sub sample of the background locations to train the model and we use the full dataset to compute the correlation between the environmental variables. We extract the sub sample using the `getSubsample` function and after we plot the background locations:
```{r, fig.align="center"}
bg_model <- getSubsample(bg, 5000, seed = 2530)
ggplot(map_data('world'), aes(long, lat)) +
    geom_polygon(aes(group = group), fill = "grey95", color = "gray40", size = 0.2) +
    geom_jitter(data = bg_model@coords, aes(x = X, y = Y),
                color = 'blue', alpha = 0.4, size = 0.5) +
    labs(x = "longitude", y = "latitude") +
    theme_minimal() +
    theme(legend.position = "none") +
    coord_fixed() +
    scale_x_continuous(limits = c(-125, -32)) +
    scale_y_continuous(limits = c(-56, 40))
```
<div style="text-align: right"><a href="#top">Back to top</a></div>

# Train MaxEnt models

The `trainMaxent` function helps training MaxEnt models in R and it gives the possibility to set several arguments. By default the function saves the **MaxEnt** output files in a temporary folder that is sistematically deleted at the end of the script but we can use the parameter `folder` to specify where to permanently save the output. The next line of code is an example of how to use the function to train a model (note that `rm = 1`, `fc = "lqph"` and `type = "cloglog"` are the default settings of MaxEnt software v. 1.4.1 when you have at least 80 presence locations):

```{r run MaxEnt}
full_model <- trainMaxent(presence = train, bg = bg_model, rm = 1, fc = "lqph", type = "cloglog", test = val, folder = "full_model")
```

The main arguments have a dedicated parameter in the function, for all the others we can use the `extra_args` parameter and pass a vector with the options (e.g. `extra_args = c("responsecurves=true", "writeplotdata=true"` to plot the response curves in the html file). The function takes care of the environmental variables data type: when we prepared the data we defined **biome** as categorical variable and the information is passed to the `trainMaxent` function.

The result of the `trainMaxent` function is an object of class `Maxent` that stores in different slots the **presence**, **background** and, if provided, the **test** dataset; MaxEnt **results**, **lambdas** coefficients, **rm**, **fc**, **iterations**, **output type** and the **folder** where is stored the MaxEnt output if provided. We can access this values using `model@name_of_the_slot`, like `model@rm` or `model@fc`. When we call the model object R prints all the parameters we used to train the model and open the **html** file in the browser ( only if we save the results permanetly using the parameter `folder = my_path`).
```{r}
full_model
```
<div style="text-align: right"><a href="#top">Back to top</a></div>

## Variable Importance

We can see the **percent contribution** and the **permutation importance** of each environmental variable using `varImp`:

```{r variable contribution}
varImp(full_model)
```

Or plot it using the `plotVarImp` function (the default color is grey):

```{r plot model, fig.align="center"}
plotVarImp(full_model, type = "perm", color = '#159957')
```

Another way to estimate the variable importance is to run a **Jackknife Test**:
```{r jk, eval=F}
jk <- doJk(full_model, with_only = T, metric = "auc")
```

```{r, echo=F}
jk <- suppressMessages(doJk(full_model, with_only = T, metric = "auc"))
```
If the model has a test dataset, it returns the train and test AUC, if not it returns only the train AUC
```{r show jk}
jk
```
We can plot the result of the **Jackknife Test** with the `plotJk` function:
```{r plot jk, fig.align="center"}
plotJk(jk, type = "train", ref = auc(full_model))
```
The red dashed vertical line represent the AUC value of the model trained using all the environmental variabled.
<div style="text-align: right"><a href="#top">Back to top</a></div>

## Plot Response Curves

The function `plotResponse` plots the response curve of the given environmental variable:

```{r bio12, fig.align="center"}
plotResponse(full_model, variable = 'bio1', marginal = T, fun = mean, rug = T, clamp = T)
```

**bio1** is a continuous variable, **biome** is categorical (in the next graph we plot the univariate response curve):

```{r biome, fig.align="center"}
plotResponse(full_model, variable = 'biome', marginal = F, color = '#159957', clamp = T)
```

`color` allow to change the **color** of the plot and `marginal` if you want to plot the marginal response curve.
<div style="text-align: right"><a href="#top">Back to top</a></div>

## Plot ROC curve

```{r plot ROC, fig.align="center"}
plotROC(full_model, val = val)
```
<div style="text-align: right"><a href="#top">Back to top</a></div>

# Make Prediction

The function `trainMaxent` trains a model but doesn't make any prediction. The `predict` method makes the prediction:
```{r make prediction}
prediction <- predict(full_model, data = predictors, filename = "full_model_prediction", format = "GTiff", overwrite = T)
```
You can pass as `data` a `data frame`, a `SWD object` or a [Raster Stack](https://www.rdocumentation.org/packages/raster/versions/2.6-7/topics/stack), [Raster Brick](https://www.rdocumentation.org/packages/raster/versions/2.6-7/topics/brick) object. In the first case the function returns a vector of prediction, in the second case it returns a raster map that can be saved providing the `filename` parameter. For big dataset you can provide the parameter `progress = "text"` to display a progress bar and take advantage of the parallel computation using `parallel = T`.

## Plot Prediction

The prediction can be plotted using any of the `plot` function in R but the function `plotPred` provides a way to plot it using the [rasterVis](https://oscarperpinan.github.io/rastervis/) package:
```{r plot prediction, fig.align="center"}
plotPred(prediction, lt = "cloglog output")
```
The default colorramp is similar to the usual MaxEnt output, but you can also use different colorramps:

```{r plot prediction colorramp, fig.align="center"}
plotPred(prediction, colorramp = c("#2c7bb6", "#abd9e9", "#ffffbf", "#fdae61", "#d7191c"), lt = "cloglog output")
```

To get nice combination of colors for maps visit [COLORBREWER](http://colorbrewer2.org/), the one in this example is the **5-class RdYlBu** colorblind safe and print friendly.
Using `hr = T` you obtain a high resolution plot, useful for publications.
<div style="text-align: right"><a href="#top">Back to top</a></div>

## Plot Presence Absence

To create a presence absence map we choose a threshold provided by MaxEnt. The function `maxentTh` provides a list of possible thresholds:

```{r threshold, eval=F}
maxentTh(full_model)
```
```{r, echo=F}
kable(maxentTh(full_model)) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "responsive")) %>%
  scroll_box(height = "500px")
```

For example we use the **Maximum training sensitivity plus specificity Cloglog threshold** as parameter for the `plotPA` function:

```{r plot presence absence, fig.align="center"}
plotPA(prediction, th = 0.0476, filename = 'pa_map', overwrite = TRUE, format = 'GTiff')
```

or we can customise the colors using:

```{r plot presence absence colors, fig.align="center"}
plotPA(prediction, th = 0.2849, filename = 'pa_map', overwrite = TRUE, colors = c("#7fbf7b", "#af8dc3"))
```
<div style="text-align: right"><a href="#top">Back to top</a></div>

# Data Driven Variable Selection

Some of the environmental variables used to train our **full_model** are highly correlated.
There are two functions that help to visualize if the environmental variables are highly correlated: `plotCor` and `corVar`.
`plotCor` builds a correlation matrix heat map (we use all the extracted background points to checks for autocorrelation):
```{r heat map, fig.align="center"}
plotCor(bg, method = "spearman", cor_th = 0.75)
```
Using the parameter `correlationth` the function prints only the value that are higher than the absolute value of the given threshold (in this case 0.75).
`corVar` creates a data frame with the correlated variables:
```{r varCor}
corVar(bg, method = "spearman", cor_th = 0.75)
```

One possible approach to remove highly correlated variables is to choose a subset of variables based on expert knowledge of the target species. If you don't know which variable to retain, you can use some kind of data-driven variable selection. The function `varSel` iterates through several steps: starting from a trained model it checks if the variable ranked as most important (both variable importance ranks are possible: percent contribution and permutation importance) is highly correlated with other variables according to the given method and correlation threshold. In this case it performs a Jaccknife test and among the correlated variables it discards the one that decreases the less the model performance. After it trains another model and checks again for highly correlated variables. The process is repeated until the retained variables are not highly correlated anymore. We suggest to use a very low regularization parameter to perform variable selection, in this case we use 0.001 (see Vignali et al. we add here the publication)
```{r variable selection, eval=F}
selected_var_model <- varSel(full_model, bg, metric = "auc", rm = 0.001, method = "spearman", cor_th = 0.75, use_permutation = T)
```
```{r, echo=F}
selected_var_model <- suppressMessages(varSel(full_model, bg, metric = "auc", rm = 0.001, method = "spearman", cor_th = 0.75, use_permutation = T))
```
The function returns the model trained without the correlated variables, if we print the object we see that the function removed **bio1**, **bio16**, **bio7** and **bio8**:
```{r}
selected_var_model
```
If we plot the ROC curve we see that after removing the correlated variables the model has still very good performances:
```{r}
plotROC(selected_var_model, val = val)
```
<div style="text-align: right"><a href="#top">Back to top</a></div>

# Tune number of background locations
