---
title: "Evaluate a model"
bibliography: ../SDMtune.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(knitr.table.format = "html")
```

```{r load data, echo=FALSE, message=FALSE}
default_model <- SDMtune:::bm_maxent
files <- list.files(path = file.path(system.file(package = "dismo"), "ex"), pattern = "grd", full.names = TRUE)
predictors <- raster::stack(files)
```

## Intro

In the previous articles you have learned how to [prepare the data](./prepare-data.html) for the analysis, how to [train a model](./train-model.html) and how to [make predictions](./make-predictions.html) using **SDMtune**. In this article you will learn how to evaluate your model using three different metrics.

`SDMtune` implements three evaluation metrics:

* AUC: Area Under the ROC curve [@Fielding1997]
* TSS: True Skill Statistic [@ALLOUCHE2006]
* AICc: Akaike Information Criterion corrected for small sample size [@Warren2011a]

We will compute the value of the metrics on the training dataset, using the `default model` that we trained in a previous article.

## AUC

As usually we first load the **SDMtune** package:

```{r load SDMtune, warning=FALSE}
library(SDMtune)
```

The AUC can be calculated using the function `auc()`:
```{r auc}
auc(default_model)
```

We can also plot the ROC curve using the function `plotROC()`:
```{r plot roc, fig.align="center"}
plotROC(default_model)
```

## TSS

The TSS is computed with the function `tss()`:
```{r tss}
tss(default_model)
```

## AICc

For the AICc we use the function `aicc()`. In this case we need to pass to the `env` argument the 'predictors' raster `stack` object that we created in the [first article](./prepare-data.html#acquire-environmental-variables):
```{r aicc}
aicc(default_model, env = predictors)
```

### Try yourself

Try to compute the three metrics using a model trained using the **Maxnet** method.

## Conclusion

In this article you have learned:

* how to calculate the AUC;
* how to plot the ROC curve;
* how to calculate the TSS;
* how to calculate the AICc.

In the [next article](./evaluation-strategies.html) you will learn two different strategies that can be used to correctly evaluate the model performance.

### References
