---
title: "Prepare data for the analysis"
bibliography: ../SDMtune.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE,
                      comment = "#>",
                      fig.align = "center")
options(knitr.table.format = "html")
```

## Intro

In this article you will learn how to prepare the data to train models using `SDMtune`. We will use the [virtualSp](../../reference/virtualSp.html) dataset included in the package and environmental predictors from the  [WorldClim](https://www.worldclim.org/) dataset.

## Set working environment

Load the required packages for the analysis:

```{r load-pkgs, message=FALSE, warning=FALSE}
library(ggplot2)    # To plot locations
library(maps)       # To access useful maps
library(rasterVis)  # To plot raster objects
```

## Acquire environmental variables

For the analysis we use the climate data of [WorldClim](https://www.worldclim.org/) version 1.4 [@Hijmans2005] and the [terrestrial ecoregions](https://www.worldwildlife.org/publications/terrestrial-ecoregions-of-the-world)  from WWF [@Olson2001] included in the `dismo` package:

```{r get-predictors}
files <- list.files(path = file.path(system.file(package = "dismo"), "ex"),
                    pattern = "grd",
                    full.names = TRUE)
```

We convert the files in a [raster](https://www.rdocumentation.org/packages/terra/versions/1.6-17/topics/rast) object that will be used later in the analysis:

```{r create-raster-object}
predictors <- terra::rast(files)
```

There are nine environmental variables, eight continuous and one categorical:

```{r names-predicrtors}
names(predictors)
```

* Continuous variables
    + **bio1** Annual Mean Temperature
    + **bio5** Max Temperature of Warmest Month
    + **bio6** Min Temperature of Coldest Month
    + **bio7** Temperature Annual Range (bio5-bio6)
    + **bio8** Mean Temperature of Wettest Quarter
    + **bio12** Annual Precipitation
    + **bio16** Precipitation of Wettest Quarter
    + **bio17** Precipitation of Driest Quarter
* Categorical variables
    + **biome** Terrestrial Ecoregions of the World

We can plot **bio1** using the [gplot](https://www.rdocumentation.org/packages/rasterVis/versions/0.43/topics/gplot-methods) function from the `rasterVis` package:

```{r plot-bio1}
gplot(predictors$bio1) +
    geom_tile(mapping = aes(fill = value)) +
    coord_equal() +
    scale_fill_gradientn(colours = c("#2c7bb6", "#abd9e9", "#ffffbf", "#fdae61", "#d7191c"),
                         na.value = "transparent",
                         name = "°C x 10") +
    labs(title = "Annual Mean Temperature",
         x = "longitude",
         y = "latitude") +
    scale_x_continuous(expand = c(0, 0)) +
    scale_y_continuous(expand = c(0, 0)) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5),
          axis.ticks.x = element_blank(),
          axis.ticks.y = element_blank())
```

## Prepare presence and background locations

Let's load the **SDMtune** package:

```{r load-SDMtune}
library(SDMtune)
```

For demonstrating how to use `SDMtune` we use the random generated virtual species [virtualSp](../../reference/virtualSp.html) dataset included in the package. The dataset contains `r nrow(virtualSp$presence)` coordinates for presence and `r nrow(virtualSp$background)` for background locations.

```{r load-data}
help(virtualSp)
p_coords <- virtualSp$presence
bg_coords <- virtualSp$background
```

Plot the study area together with the presence locations:

```{r plot-presence}
ggplot(data = map_data("world"), mapping = aes(x = long, y = lat)) +
    geom_polygon(aes(group = group), fill = "grey95", color = "gray40", size = 0.2) +
    geom_jitter(data = p_coords, aes(x = x, y = y), color = "red",
                alpha = 0.4, size = 1) +
    labs(x = "longitude", y = "latitude") +
    theme_minimal() +
    theme(legend.position = "none") +
    coord_fixed() +
    scale_x_continuous(limits = c(-125, -32)) +
    scale_y_continuous(limits = c(-56, 40))
```

To plot the background locations run the following code:

```{r plot-bg_model-locations}
ggplot(data = map_data("world"), mapping = aes(x = long, y = lat)) +
    geom_polygon(aes(group = group), fill = "grey95", color = "gray40", size = 0.2) +
    geom_jitter(data = as.data.frame(bg_coords), aes(x = x, y = y),
                color = "blue", alpha = 0.4, size = 0.5) +
    labs(x = "longitude", y = "latitude") +
    theme_minimal() +
    theme(legend.position = "none") +
    coord_fixed() +
    scale_x_continuous(limits = c(-125, -32)) +
    scale_y_continuous(limits = c(-56, 40))
```

## Create an SWD object

Before training a model we have to prepare the data in the correct format. The `prepareSWD()` function creates an `SWD()` object that stores the species name, the coordinates of the species at presence and absence/background locations and the value of the environmental variables at the locations. The argument `categorical` indicates which environmental variables are categorical. In our example **biome** is categorical (we can pass a vector if we have more than one categorical environmental variable). The function extracts the value of the environmental variables for each location and excludes those locations that have `NA` value for at least one environmental variable.

```{r prepare-SWD, eval=FALSE}
data <- prepareSWD(species = "Virtual species",
                   p = p_coords,
                   a = bg_coords,
                   env = predictors,
                   categorical = "biome")
```
```{r prepare-SWD-output, echo=FALSE}
data <- prepareSWD(species = "Virtual species",
                   p = p_coords,
                   a = bg_coords,
                   env = predictors,
                   categorical = "biome",
                   verbose = FALSE)
```

## Explore the SWD object

Let's have a look at the created `SWD()` object:

```{r show-SWD-object}
data
```

When we print an `SWD()` object we get a bunch of information:

* the name of the class;
* the name of the species;
* the number of presence locations;
* the number of absence/background locations;
* the environmental variables available in the dataset:
    + the name of the continuous environmental variables, if any;
    + the name of the categorical environmental variables, if any.

The object contains four slots: `@species`, `@coords` `@data` and `@pa`. `@pa` contains a vector with 1 for presence and 0 for absence/background locations. To visualize the data we run:

```{r show-data, eval=FALSE}
head(data@data)
```
```{r show data output, echo=FALSE}
kableExtra::kable(head(data@data)) |> 
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                            position = "center",
                            full_width = FALSE)
```

We can visualize the coordinates with:

```{r show-coords-data, eval=FALSE}
head(data@coords)
```
```{r show coords data output, echo=FALSE}
kableExtra::kable(head(data@coords)) |> 
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                            position = "center",
                            full_width = FALSE)
```

or the name of the species with:

```{r show-species}
data@species
```

## Save an SWD object

We can save the `SWD()` object in a **.csv** file using the function `swd2csv()` (the function saves the file in the working directory). There are two possibilities:

* save the object in a single file with the column **pa** indicating if the location is a presence (1) or an absence/background (0) site

```{r save-SWD-single, eval=FALSE}
swd2csv(data, file_name = "data.csv")
```

* save the object in two separate files: one for the presence and the other for the absence/background locations

```{r save-SWDdouble, eval=FALSE}
swd2csv(data, file_name = c("presence.csv", "background.csv"))
```

## Conclusion

In this article you have learned:

* how to create a raster object;
* how to plot a raster object using the `gplot` function included in the `rasterVis` package;
* how to plot locations using the `ggplot` and the `maps` packages;
* how to create an `SWD()` objects;
* how to extract information from an `SWD()` object;
* how to save an `SWD()` object in a **.csv** file.

Move on to the [second article](./train-model.html) and learn how to train models using `SDMtune`.

# References
