---
title: "Prepare data for the analysis"
bibliography: ../SDMtune.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(knitr.table.format = "html")
```

## Intro

In this article you will learn how to prepare the data to train models using `SDMtune`. We will use the [condor](../../reference/condor.html) dataset included in the package and environmental predictors from the  [WorldClim](http://www.worldclim.org/) dataset.

## Set working environment

Load the required packages for the analysis:

```{r load pkgs, message=FALSE, warning=FALSE}
library(ggplot2)    # To plot locations
library(maps)       # To access useful maps
library(rasterVis)  # To plot raster objects
```

## Acquire environmental variables

For the analysis we use the climate data of [WorldClim](http://www.worldclim.org/) version 1.4 [@Hijmans2005] and the [terrestrial ecoregions](https://www.worldwildlife.org/publications/terrestrial-ecoregions-of-the-world)  from WWF [@Olson2001] included in the `dismo` package:

```{r get predictors, fig.align="center"}
files <- list.files(path = file.path(system.file(package = "dismo"), "ex"), pattern = "grd", full.names = TRUE)
```

We convert the files in a [raster stack](https://www.rdocumentation.org/packages/raster/versions/2.8-19/topics/stack) object that will be used later in the analysis:
```{r create raster stack}
predictors <- stack(files)
```

There are nine environmental variables, eight continuous and one categorical:
```{r names predicrtors}
names(predictors)
```

* Continuous variables
    + **bio1** Annual Mean Temperature
    + **bio5** Max Temperature of Warmest Month
    + **bio6** Min Temperature of Coldest Month
    + **bio7** Temperature Annual Range (bio5-bio6)
    + **bio8** Mean Temperature of Wettest Quarter
    + **bio12** Annual Precipitation
    + **bio16** Precipitation of Wettest Quarter
    + **bio17** Precipitation of Driest Quarter
* Categorical variables
    + **biome** Terrestrial Ecoregions of the World

We can plot **bio1** using the [gplot](https://www.rdocumentation.org/packages/rasterVis/versions/0.43/topics/gplot-methods) function from the `rasterVis` package:

```{r plot bio1, fig.align="center"}
gplot(predictors$bio1) +
    geom_tile(aes(fill = value)) +
    coord_equal() +
    scale_fill_gradientn(colours = c("#2c7bb6", "#abd9e9", "#ffffbf", "#fdae61", "#d7191c"),
                         na.value = "transparent",
                         name = "°C x 10") +
    labs(title = "Annual Mean Temperature",
         x = "longitude",
         y = "latitude") +
    scale_x_continuous(expand = c(0, 0)) +
    scale_y_continuous(expand = c(0, 0)) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5),
          axis.ticks.x = element_blank(),
          axis.ticks.y = element_blank())
```

## Prepare presence locations

Let's load the **SDMtune** package:

```{r load SDMtune, warning=FALSE}
library(SDMtune)
```

For demonstrating how to use `SDMtune` we use the [condor](../../reference/condor.html) dataset included in the package. The dataset contains `r nrow(condor)` locations of the Andean condor [*Vultur gryphus*](https://en.wikipedia.org/wiki/Andean_condor) downloaded from the Global Biodiversity Inventory Facility [GBIF](http://www.gbif.org/).
```{r preview condor}
help(condor)
head(condor)
```

We select the the first two columns that contain the coordinates of the locations:
```{r prepare presence}
p_coords <- condor[, 1:2]
```

Plot the study area together with the presence locations:
```{r plot presence, fig.align="center"}
ggplot(map_data("world"), aes(long, lat)) +
    geom_polygon(aes(group = group), fill = "grey95", color = "gray40", size = 0.2) +
    geom_jitter(data = p_coords, aes(x = x, y = y), color = "red",
                alpha = 0.4, size = 1) +
    labs(x = "longitude", y = "latitude") +
    theme_minimal() +
    theme(legend.position = "none") +
    coord_fixed() +
    scale_x_continuous(limits = c(-125, -32)) +
    scale_y_continuous(limits = c(-56, 40))
```
In this tutorial we don't check the validity of the locations, instead we assume they are correct and we model the distribution only for demonstration purpose.

## Prepare background locations

We extract 10000 background locations (default number in the MaxEnt software ) using the [randomPoints](https://www.rdocumentation.org/packages/dismo/versions/1.1-4/topics/randomPoints) function included in the `dismo` package (we set the seed to have reproducible results).

```{r prepare background}
set.seed(25)
bg_coords <- dismo::randomPoints(predictors, 10000)
```

The environmental variables we downloaded have a coarse resolution and the function can extract only `r nrow(bg_coords)` instead of 10000 background locations (see the warning message).

## Create an SWD object

Before training a model we have to prepare the data in the correct format. The `prepareSWD()` function creates an `SWD()` object that stores the species name, the coordinates of the species locations and the value of the environmental variables at the locations. The argument `categorical` indicates which environmental variables are categorical. In our example **biome** is categorical (we can pass a vector if we have more than one categorical environmental variable). The function extracts the value of the environmental variables for each location and excludes those locations that have `NA` value for at least one environmental variable.

```{r prepare presence SWD}
presence <- prepareSWD(species = "Vultur gryphus", coords = p_coords, env = predictors, categorical = "biome")

```
```{r prepare bg SWD}
bg <- prepareSWD(species = "Vultur gryphus", coords = bg_coords, env = predictors, categorical = "biome")
```

The warning message reports that in the background dataset `r nrow(bg_coords) - nrow(bg@data)`  locations were discarded. The final number of background locations is `r nrow(bg@data)`.

## Explore an SWD object

Let's have a look at the created `SWD()` object:

```{r show SWD object}
presence
```

When we print an `SWD()` object we get a bunch of information:

* the name of the class;
* the name of the species;
* the number of locations;
* the environmental variables available in the dataset:
    + the name of the continuous environmental variables, if any;
    + the name of the categorical environmental variables, if any.

The object contains three slots: `species`, `coords` and `data`. To visualize the data we run:

```{r show presence data}
head(presence@data)
```

We can visualize the coordinates with:

```{r show presence coords data}
head(presence@coords)
```

or the name of the species with:
```{r show presence species}
presence@species
```

We can save an `SWD()` object in a **.csv** file using the function `swd2csv()` (the function saves the file in the working directory):
```{r save SWD, eval=F}
swd2csv(presence, file_name = "presence.csv")
```

## Subset the background locations:

We will use a sub sample of the background locations to train the model and the full dataset to explore the correlation between the environmental variables. We extract the sub sample using the function `getSubsample()`:

```{r get bg subsample}
bg_model <- getSubsample(bg, 5000, seed = 25)
```

To plot the retained background locations run the following code:

```{r plot bg_model locations, fig.align="center"}
ggplot(map_data("world"), aes(long, lat)) +
    geom_polygon(aes(group = group), fill = "grey95", color = "gray40", size = 0.2) +
    geom_jitter(data = bg_model@coords, aes(x = X, y = Y),
                color = "blue", alpha = 0.4, size = 0.5) +
    labs(x = "longitude", y = "latitude") +
    theme_minimal() +
    theme(legend.position = "none") +
    coord_fixed() +
    scale_x_continuous(limits = c(-125, -32)) +
    scale_y_continuous(limits = c(-56, 40))
```

## Conclusion

In this article you have learned:

* how to create a raster `stack` object;
* how to plot a raster object using the `gplot` function included in the `rasterVis` package;
* how to create an `SWD()` objects;
* how to extract information from an `SWD()` object;
* how to save an `SWD()` object in a **.csv** file;
* how to extract a sub sample of an `SWD()` object;
* how to plot locations using the `ggplot` and the `maps` packages.

Move on to the [second article](./train_model.html) and learn how to train models using `SDMtune`.

### References
