---
title: "Train a model"
bibliography: ../SDMtune.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE,
                      comment = "#>",
                      fig.align = "center")
options(knitr.table.format = "html")
```

```{r load-data, echo=FALSE, message=FALSE}
default_model <- SDMtune:::bm_maxent
```

## Intro

In the [previous article](./prepare-data.html) you have learned how to prepare the data for the analysis using the [virtualSp](../../reference/virtualSp.html) dataset and the [WorldClim](https://www.worldclim.org/) environmental variables. Now it's time to train your first model, let's do it!  

`SDMtune` supports four methods for model training:

* Artificial Neural Networks **ANN**, using the `nnet` package [@Venables2002];
* Boosted Regression Trees **BRT**, using the `gbm` package [@Greenwell2019];
* Maximum Entropy with two implementations:
    * **Maxent** using the `dismo` package [@Hijmans2017];
    * **Maxnet** using the `maxnet` package [@Phillips2017b];
* Random Forest **RF**, using the `randomForest` package [@Liaw2002].

The code necessary to train a model is the same for all the implementations. We will show how to train a **Maxent** model, you can adapt the code for the other methods or check this [article](./train-tune-presence-absence-models.html).

## Train a model with default settings

First we load the **SDMtune** package:

```{r load-SDMtune}
library(SDMtune)
```


We use the function `train()` to train a **Maxent** model. We need to provide two arguments:

* `method`: "Maxent" in our case;
* `data`: the `SWD()` object with the presence and background locations that we created in the previous article.

```{r default-model, echo=FALSE}
default_model <- SDMtune:::bm_maxent
```
```{r train, eval=FALSE}
default_model <- train(method = "Maxent",
                       data = data)
```

The function trains the model using default settings that are:

* linear, quadratic, product and hinge feature class combinations;
* regularization multiplier equal to 1;
* 500 algorithm iterations.

We will see later how to change the default settings, for the moment let's have a look at the `default_model` object.

## Explore an SDMmodel object

The output of the function `train()` is an object of class `SDMmodel()`. Let's print it:

```{r print-model}
default_model
```

When we print an `SDMmodel` object we get the following information:

* the name of the class;
* the method used to train the model;
* the name of the species;
* the number of presence locations;
* the number of absence/background locations;
* the model configurations:
    * fc: the feature class combinations;
    * reg: the regularization multiplier;
    * iter: the number of iterations;
* the environmental variables used to train the model:
    * the name of the continuous environmental variables, if any;
    * the name of the categorical environmental variables, if any.
    
An `SDMmodel()` object has two slots:

```{r get-slots}
slotNames(default_model)
```

* **data**: an `SWD()` object with the presence absence/background locations used to train the model;
* **model**: a `Maxent()` object, in our case, with all the model configurations.

The slot `model` contains the configurations of the model plus other information used to make predictions.

```{r model-slots}
slotNames(default_model@model)
```

For the moment the most important are: **fc**, **reg** and **iter** that contain the values of the model configuration. We will explore the others later in another article.

## Train a model changing the default settings

The function `train()` accepts optional arguments that can be used to change the default model settings. In our previous example we could have trained the same model using:

```{r retrain, eval=FALSE}
default_model <- train(method = "Maxent", 
                       data = data, 
                       fc = "lqph", 
                       reg = 1, 
                       iter = 500)
```

### Try yourself

Try to change the default settings  and train a model using linear and hinge as feature class combination, 0.5 as regularization multiplier and 700 iterations. To see the solution highlight the next cell:

```{r exercise1, eval=FALSE, class.source='exercise'}
model <- train(method = "Maxent", 
               data = data, 
               fc = "lh", 
               reg = 0.5, 
               iter = 700)
```

By default Maxent models are trained using the arguments *"removeduplicates=false"* and *"addsamplestobackground=false"*. The user should have the full control of the data used to train the model, so is expected that duplicated locations are already removed and that the presence locations are already included in the background locations, when desired. You can use the function `thinData()` to remove duplicated locations and the function `addSamplesToBg()` to add the presence locations to the background locations.

## Train a Maxnet model

Train a model using the **Maxnet** method is as simple as changing the name of the method in the `train()` function, the only difference here is that we cannot set the number of iteration.

### Try yourself

Try to train a model using the **Maxnet** method. To see the solution highlight the following cell:

```{r train-maxnet, eval=FALSE, class.source='exercise'}
maxnet_model <- train("Maxnet", 
                      data = data)
```

## Conclusion

In this article you have learned:

* how to train a **Maxent** model using default settings;
* how to explore an `SDMmodel()` object;
* how to train a model changing the default settings;
* how to train a model using the **Maxnet** method.

In the [next article](./make-predictions.html) you will learn how to use the model that you have just trained to get the predicted value for new localities.

### References
