---
title: "Train a model"
bibliography: ../SDMtune.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(knitr.table.format = "html")
```

```{r load data, echo=FALSE, message=FALSE}
default_model <- SDMtune:::bm_maxent
```

## Intro

In the [previous article](./prepare_data.html) you have learned how to prepare the data for the analysis using the [condor](../../reference/condor.html) dataset and the [WorldClim](http://www.worldclim.org/) environmental variables. Now it's time to train your first model, let's do it!  
**SDMtune** at the moment supports only the Maximum Entropy method with two different implementations:

* **Maxent** using the `dismo` package [@Hijmans2017];
* **Maxnet** using the `maxnet` package [@Phillips2017a].

The code necessary to train a model is the same for both implementations.

## Check that MaxEnt is correctly configured

Before training a model we check that MaxEnt is correctly configured for the `dismo` package:
```{r check maxent, eval=FALSE}
dismo::maxent()
```

In case you get an error, check the [Get Started](../SDMtune.html) page to see how to configure MaxEnt.

## Train a model with default settings

First we load the **SDMtune** package:

```{r load SDMtune, warning=FALSE}
library(SDMtune)
```


We use the function `train()` to train a **Maxent** model. We need to provide three arguments:

* `method`: "Maxent" in our case;
* `p`: the `presence` `SWD()` object we created in the previous article;
* `a`: the `bg_model` `SWD()` object we created in the previous article as sub sample of the all extracted background locations.

```{r, echo=FALSE}
default_model <- SDMtune:::bm_maxent
```
```{r train, eval=FALSE}
default_model <- train(method = "Maxent", p = presence, a = bg_model)
```

The function trains the model using default settings that are:

* linear, quadratic, product and hinge feature class combinations;
* regularization multiplier equal to 1;
* 500 algorithm iterations.

We will see later how to change the default settings, for the moment let's have a look at the `default_model` object.

## Explore an SDMmodel object

The output of the `train()` function is an object of class `SDMmodel()`. Let's print the `model` object we've just created:
```{r print model}
default_model
```

When we print an `SDMmodel` object we get the following information:

* the name of the class;
* the method used to train the model;
* the name of the species;
* the number of presence locations;
* the model configurations:
    + a: the number of background locations;
    + fc: the feature class combinations;
    + reg: the regularization multiplier;
    + iter: the number of iterations;
* the environmental variables used to train the model:
    + the name of the continuous environmental variables, if any;
    + the name of the categorical environmental variables, if any.
    
An `SDMmodel()` object has three slots:

```{r get slots}
slotNames(default_model)
```

* **p**: an `SWD()` object with the presence locations used to train the model;
* **a**: an `SWD()` object with the background locations used to train the model;
* **model**: a `Maxent()` object, in our case, with all the model configurations.

We can access the different slots in the same way we explored the `SWD()` object in the previous article, using the **\@** symbol before the slot name (e.g. `default_model@p`).

The slot `model` contains the configurations of the model plus other information used to make predictions.

```{r model slots}
slotNames(default_model@model)
```

For the moment the most important are: **fc**, **reg** and **iter** that contain the values of the model configuration. We will explore the others later in another article.

## Train a model changing the default settings

The function `train()` accepts optional arguments that can be used to change the default model settings. In our previous example we could have trained the model also using:

```{r retrain, eval = FALSE}
default_model <- train(method = "Maxent", p = presence, a = bg_model, fc = "lqph", reg = 1, iter = 500)
```

### Try yourself
Try to change the default settings  and train a model using linear and hinge as feature class combination, 0.5 as regularization multiplier and 700 iterations. To see the solution highlight the next cell:

```{r exercise1, eval = FALSE, class.source='exercise'}
model <- train(method = "Maxent", p = presence, a = bg_model, fc = "lh", reg = 0.5, iter = 700)
```

## Train a Maxnet model

Train a model using the **Maxnet** method is as simple as changing the name of the method in the `train()` function, the only difference here is that we cannot set the number of iteration.

### Try yourself

Try to train a model using the **Maxnet** method. To see the solution highlight the following cell:

```{r train maxnet, eval = FALSE, class.source='exercise'}
maxnet_model <- train("Maxnet", p = presence, a = bg_model)
```

## Conclusion

In this article you have learned:

* how to train a **Maxent** model using default settings;
* how to explore an `SDMmodel()` object;
* how to train a model changing the default settings;
* how to train a model using the **Maxnet** method.

In the [next article](./make_predictions.html) you will learn how to use the model that you have just trained to get the predicted value for new localities.

### References
