---
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
params:
  model: model
  type: type
  folder: folder
  test: test
  response_curves: true
  jk: false
---

```{r, echo=F}
library(SDMsel)
library(kableExtra)
```

**`r class(params$model@model)`** output produced using **SDMsel** version `r packageVersion("SDMsel")` (Vignali S. *et al.*, 2018).

```{r, echo=FALSE, message=FALSE}
message("save files...")
swd2csv(params$model@presence, paste0(params$folder, "/presence.csv"))
swd2csv(params$model@background, paste0(params$folder, "/background.csv"))
if (!is.null(params$test))
  swd2csv(params$test, paste0(params$folder, "/test.csv"))
plot_folder <- paste0(params$folder, "/plots")
```


# ROC curve

Next we plot the ROC curve for `r model@presence@species`:

```{r plot ROC, echo=FALSE, fig.align="center", message=FALSE}
message("plotting ROC curve...")
plot <- plotROC(params$model, test = params$test)
suppressMessages(ggsave(filename = "ROC_curve.png", plot = plot, device = "png", path = plot_folder))
element <- "<p><a href=\"./plots/ROC_curve.png\"><img src=\"./plots/ROC_curve.png\" style=\"width: 70%; display: block; margin-left: auto; margin-right: auto;\"></a></p>"
htmltools::HTML(element)
```

# Thresholds

```{r, echo=FALSE, message=FALSE}
message("doing thresholds...")
knitr::kable(thresholds(params$model, type = params$type, test = params$test), digits = 20) %>%
  kable_styling(bootstrap_options = c("responsive"))
```

`r if(!params$response_curves) {"<!--"}`
# Response curves

Marginal curves:

```{r, fig.hold='hold', out.width='25%', message=FALSE, echo=FALSE}
if (params$response_curves) {
  vars <- sort(colnames(params$model@presence@data))
  pb <- progress::progress_bar$new(
    format = "Marginal curves [:bar] :percent in :elapsedfull", total = length(vars), clear = FALSE, width = 60, show_after = 0)
  pb$tick(0)
  elements <- c("<p>")
  for (var in vars) {
    plot <- suppressMessages(plotResponse(params$model, var, type = params$type, marginal = TRUE, rug = TRUE))
    suppressMessages(ggsave(filename = paste0(var, "_marginal.png"), plot = plot, device = "png", path = plot_folder))
    path <- paste0(plot_folder, "/", var, "_marginal.png")
    element <- paste0("<a href=\"", path, "\"><img src=\"", path, "\" width=25% title=\"", var, "\"></a>")
    elements <- c(elements, element)
    pb$tick(1)
  }
  elements <- c(elements, "</p>")
  elements <- paste(elements, collapse = "")
  htmltools::HTML(elements)
}
```

Univariate curves:

```{r, message=FALSE, echo=FALSE}
if (params$response_curves) {
  pb <- progress::progress_bar$new(
    format = "Univariate curves [:bar] :percent in :elapsedfull", total = length(vars), clear = FALSE, width = 60, show_after = 0)
  pb$tick(0)
  elements <- c("<p>")
  for (var in vars) {
    plot <- suppressMessages(plotResponse(params$model, var, type = params$type, marginal = FALSE, rug = TRUE))
    suppressMessages(ggsave(filename = paste0(var, ".png"), plot = plot, device = "png", path = plot_folder))
    path <- paste0(plot_folder, "/", var, ".png")
    element <- paste0("<a href=\"", path, "\"><img src=\"", path, "\" width=25% title=\"", var, "\"></a>")
    elements <- c(elements, element)
    pb$tick(1)
  }
  elements <- c(elements, "</p>")
  elements <- paste(elements, collapse = "")
  htmltools::HTML(elements)
}
```
`r if(!params$response_curves) {"-->"}`

# Variable importance

```{r, echo=FALSE, message=FALSE}
knitr::kable(varImp(params$model)) %>%
  kable_styling(bootstrap_options = c("responsive"), full_width = FALSE)
```

`r if(!params$jk) {"<!--"}`
# Jackknife test

```{r, echo=FALSE, message=FALSE}
if (jk) {
  jk <- doJk(params$model, metric = "auc", test = params$test)
  plot <- plotJk(jk, type = "train", ref = auc(params$model))
  suppressMessages(ggsave(filename = "train_jk.png", plot = plot, device = "png", path = plot_folder))
  if (!is.null(test)) {
    plot <- plotJk(jk, type = "test", ref = auc(params$model, test))
    suppressMessages(ggsave(filename = "test_jk.png", plot = plot, device = "png", path = plot_folder))
    element <- "<p><a href=\"./plots/train_jk.png\"><img src=\"./plots/train_jk.png\" width=50%></a><a href=\"./plots/test_jk.png\"><img src=\"./plots/test_jk.png\" width=50%></a></p>"
  } else {
    element <- "<p><a href=\"./plots/train_jk.png\"><img src=\"./plots/train_jk.png\" style=\"width: 70%; display: block; margin-left: auto; margin-right: auto;\"></a></p>"
  }
  htmltools::HTML(element)
}
```
`r if(!params$jk) {"-->"}`

# Model settings

Shown below the settings used to train the model:

* Model type: `r class(params$model@model)`
* Output type: `r params$type`
* Train locations: `r nrow(params$model@presence@data)`
* Background locations: `r nrow(params$model@background@data)``r if(!is.null(params$test)) {paste("\n* Test locations:", nrow(params$test@data))}`
* Continuous variables: `r names(Filter(is.numeric, params$model@presence@data))`
* Categorical variables: `r names(Filter(is.factor, params$model@presence@data))`
* Feature combination: `r params$model@model@fc`
* Regularization multiplier: `r params$model@model@rm``r if(class(params$model@model) == "Maxent") {paste("\n* Extra arguments:", paste(params$model@model@extra_args, collapse = ", "))}`

### Reference

- Sergio Vignali, Arnaud Barras and Veronika Braunisch (2018). SDMsel: Species Distribution Model Selection. R package version 0.0.0.9000
