---
output:
  html_document:
    toc: true
    toc_float: true
    css: report.css
params:
  model: model
  type: type
  folder: folder
  test: test
  response_curves: false
  only_presence: false
  jk: false
  env: null
  clamp: true
  permut: 10
  factors: null
---

```{r setup, echo=F, message=FALSE}
library(SDMtune)
library(kableExtra)
options(crayon.enabled = TRUE)
plot_folder <- file.path(params$folder, "plots")
```

**`r class(params$model@model)`** output produced using **SDMtune** version `r packageVersion("SDMtune")` (Vignali S. *et al.*, 2020).

```{r, echo=FALSE, message=FALSE}
message("\r- Saving files...", appendLF = FALSE)
saveRDS(params$model, file = file.path(params$folder, "model.Rds"))
swd2csv(params$model@data, file.path(params$folder, "train.csv"))
if (!is.null(params$test))
  swd2csv(params$test, file.path(params$folder, "test.csv"))
message("\r", cli::col_green(cli::symbol$tick), " Saving files...")
```

# ROC curve

```{r plot ROC, echo=FALSE, fig.align="center", message=FALSE}
message("- Plotting ROC curve...", appendLF = FALSE)
plot <- plotROC(params$model, test = params$test)
suppressMessages(ggplot2::ggsave(filename = "ROC_curve.png", plot = plot, device = "png", path = plot_folder))
path <- file.path(plot_folder, "ROC_curve.png")
element <- paste0("<a href=\"", path, "\"><img src=\"", path, "\" style=\"width: 70%; display: block; margin-left: auto; margin-right: auto;\"></a>")
htmltools::HTML(element)
message("\r", cli::col_green(cli::symbol$tick), " Plotting ROC curve...")
```

# Thresholds

```{r, echo=FALSE, message=FALSE}
message("- Computing thresholds...", appendLF = FALSE)
knitr::kable(thresholds(params$model, type = params$type, test = params$test), digits = 20) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
message("\r", cli::col_green(cli::symbol$tick), " Computing thresholds...")
```

`r if(!params$response_curves) {"<!--"}`
# Response curves {#response}

### Marginal curves:

```{r, message=FALSE, echo=FALSE, warning=FALSE}
if (params$response_curves) {
  message("- Plotting marginal response curves...", appendLF = FALSE)
  vars <- sort(colnames(params$model@data@data))
  elements <- c()
  for (var in vars) {
    plot <- plotResponse(params$model, var, type = params$type, only_presence = params$only_presence, marginal = TRUE, rug = TRUE, color = "#4bc0c0")
    suppressMessages(ggplot2::ggsave(filename = paste0(var, "_marginal.png"), plot = plot, device = "png", path = plot_folder))
    path <- file.path(plot_folder, paste0(var, "_marginal.png"))
    element <- paste0("<a href=\"", path, "\"><figure><img src=\"", path, "\" title=\"", var, "\"><figcaption>", var, "</figcaption></figure></a>")
    elements <- c(elements, element)
  }
  elements <- paste(elements, collapse = "")
  htmltools::HTML(elements)
}
```

### Univariate curves:

```{r, message=FALSE, echo=FALSE, warning=FALSE}
if (params$response_curves) {
  message("\r", cli::col_green(cli::symbol$tick), " Plotting marginal response curves...")
  message("- Plotting univariate response curves...", appendLF = FALSE)
  elements <- c("")
  for (var in vars) {
    plot <- suppressMessages(plotResponse(params$model, var, type = params$type, only_presence = params$only_presence, marginal = FALSE, rug = TRUE, color = "#4bc0c0"))
    suppressMessages(ggplot2::ggsave(filename = paste0(var, ".png"), plot = plot, device = "png", path = plot_folder))
    path <- paste0(plot_folder, "/", var, ".png")
    element <- paste0("<a href=\"", path, "\"><figure><img src=\"", path, "\" title=\"", var, "\"><figcaption>", var, "</figcaption></figure></a>")
    elements <- c(elements, element)
  }
  elements <- c(elements, "<p></p>")
  elements <- paste(elements, collapse = "")
  htmltools::HTML(elements)
}
```
```{r, echo=FALSE, message=FALSE}
# Leave this chunk outside to avoid deleting html...
if (params$response_curves) {
  message("\r", cli::col_green(cli::symbol$tick), " Plotting univariate response curves...")
}
```

`r if(!params$response_curves) {"-->"}`

`r if(is.null(params$env)) {"<!--"}`
# Map

```{r, echo=FALSE, message=FALSE}
if (!is.null(env)) {
  message("- Predicting distribution map...", appendLF = FALSE)
  pred <- predict(params$model, data = params$env, type = params$type, filename = paste0(params$folder, "/map"), overwrite = TRUE, clamp = params$clamp, factors = params$factors)
  plot <- plotPred(pred, lt = params$type, hr = TRUE, colorramp = c("#2c7bb6", "#abd9e9", "#ffffbf", "#fdae61", "#d7191c"))
suppressMessages(ggplot2::ggsave(filename = "map.png", plot = plot, device = "png", path = plot_folder))
path <- file.path(plot_folder, "map.png")
element <- paste0("<a href=\"", path, "\"><img src=\"", path, "\" style=\"width: 70%; display: block; margin-left: auto; margin-right: auto;\"></a>")
htmltools::HTML(element)
}
```
```{r, echo=FALSE, message=FALSE}
if (!is.null(env))
  message("\r", cli::col_green(cli::symbol$tick), " Predicting distribution map...")
```

`r if(is.null(params$env)) {"-->"}`

# Variable importance

```{r, echo=FALSE, message=FALSE}
message("- Computing variable importance...", appendLF = FALSE)
knitr::kable(suppressMessages(varImp(params$model, params$permut))) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = FALSE)
message("\r", cli::col_green(cli::symbol$tick), " Computing variable importance...")
```

`r if(!params$jk) {"<!--"}`
# Jackknife test

```{r, echo=FALSE, message=FALSE}
if (params$jk) {
  message("- Running Jackknife test...", appendLF = FALSE)
  jk <- suppressMessages(doJk(params$model, metric = "auc", test = params$test))
  plot <- plotJk(jk, type = "train", ref = auc(params$model))
  suppressMessages(ggplot2::ggsave(filename = "train_jk.png", plot = plot, device = "png", path = plot_folder))
  if (!is.null(test)) {
    plot <- plotJk(jk, type = "test", ref = auc(params$model, test))
    suppressMessages(ggplot2::ggsave(filename = "test_jk.png", plot = plot, device = "png", path = plot_folder))
    path1 <- file.path(plot_folder, "train_jk.png")
    path2 <- file.path(plot_folder, "test_jk.png")
    element <- paste0("<a href=\"", path1, "\"><img src=\"", path1, "\" width=50%></a><a href=\"", path2, "\"><img src=\"", path2, "\" width=50%></a>")
  } else {
    element <- paste0("<a href=\"", path1, "\"><img src=\"", path1, "\" style=\"width: 70%; display: block; margin-left: auto; margin-right: auto;\"></a>")
  }
  htmltools::HTML(element)
}
```
```{r, echo=FALSE, message=FALSE}
# Leave this chunk outside to avoid deleting html...
if (params$jk) {
  message("\r", cli::col_green(cli::symbol$tick), " Running Jackknife test...")
}
```

`r if(!params$jk) {"-->"}`

# Model settings
```{r, echo=FALSE, message=FALSE}
message("- Writing model settings...", appendLF = FALSE)
```

Shown below the settings used to train the model:

* Model type: `r class(params$model@model)`
* Train locations: `r nrow(params$model@data@data)`
    * presence: `r sum(params$model@data@pa == 1)`
    * absence/background locations: `r sum(params$model@data@pa == 0)`
`r if(!is.null(params$test)){paste("\n* Test locations:", nrow(params$test@data), "\n    * presence:", sum(params$test@pa == 1), "\n    * absence/background:", sum(params$test@pa == 0))}`
* Continuous variables: `r names(Filter(is.numeric, params$model@data@data))`
* Categorical variables: `r names(Filter(is.factor, params$model@data@data))`
```{r MEs, eval=class(params$model@model) %in% c("Maxent", "Maxnet"), results="asis", message=FALSE, echo=FALSE}
cat("* Output type: ", params$type, "\n")
cat("* Feature Class combination: ", params$model@model@fc, "\n")
cat("* Regularization multiplier: ", params$model@model@reg, "\n")
cat("* Do clamping: ", params$clamp, "\n")
```
```{r Maxent, eval=class(params$model@model) == "Maxent", results="asis", message=FALSE, echo=FALSE}
cat("* Extra arguments: ", paste(params$model@model@extra_args, collapse = ", "), "\n")
```
```{r ANN, eval=class(params$model@model) == "ANN", results="asis", message=FALSE, echo=FALSE}
cat("* Size: ", params$model@model@size, "\n")
cat("* Decay: ", params$model@model@decay, "\n")
cat("* Rang: ", params$model@model@rang, "\n")
cat("* Maxit: ", params$model@model@maxit, "\n")
```
```{r BRT, eval=class(params$model@model) == "BRT", results="asis", message=FALSE, echo=FALSE}
cat("* Distribution: ", params$model@model@distribution, "\n")
cat("* Number of trees: ", params$model@model@n.trees, "\n")
cat("* Interaction depth: ", params$model@model@interaction.depth, "\n")
cat("* Shrinkage: ", params$model@model@shrinkage, "\n")
cat("* Bag fraction: ", params$model@model@bag.fraction, "\n")
```
```{r RF, eval=class(params$model@model) == "RF", results="asis", message=FALSE, echo=FALSE}
cat("* Mtry: ", params$model@model@mtry, "\n")
cat("* Number of trees: ", params$model@model@ntree, "\n")
cat("* Node size: ", params$model@model@nodesize, "\n")
```

```{r, echo=FALSE, message=FALSE}
message("\r", cli::col_green(cli::symbol$tick)," Writing model settings...")
```


# References

- Vignali, S., Barras, A. G., Arlettaz, R., Braunisch, V. (2020). *SDMtune*: An R package to tune and evaluate species distribution models. Ecology and Evolution, **10**(20), 11488–11506. <https://doi.org/10.1002/ece3.6786>
- Vignali, S., Barras, A. G., Braunisch, V. (2022). *SDMtune*: Species Distribution Model Selection. R package version `r packageVersion("SDMtune")`. `r ifelse(grepl("9000",packageVersion("SDMtune")), "<https://github.com/ConsBiol-unibern/SDMtune>", "<https://CRAN.R-project.org/package=SDMtune>")`
